---
title: "In the United States, tornadoes are the most damaging events to the health of the population and floods have the greatest economic consequences"
author: "Julio C. S. Vasconcelos"
date: "25 de abril de 2021"
output: pdf_document
---

#1.Synopsis
In this file we have the task of analyzing severe climatic events that have happened in the United States in relation to public health and the economic consequences that have occurred. The US National Oceanic and Atmospheric Administration (NOAA) storm database is used, which records the characteristics of major storms and weather events, such as when and where they occur, as well as estimates of any fatalities, injuries and material damage. The purpose of this analysis is to avoid these negative results as far as possible.


#2. Data Processing
##2.1 Data Loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}

input <- read.csv(file = "repdata_data_StormData.csv", sep = ",", 
                  stringsAsFactors = FALSE)
input$BGN_DATE <- as.Date(input$BGN_DATE, "%m/%d/%Y %H:%M:%S")

```

##2.2 Data Selection
The events in the bank started in the year 1950 and end in November 2011.

```{r}
summary(input$BGN_DATE)
```

```{r}
recentdata<-input[input$BGN_DATE>as.Date("2002-01-01", "%Y-%m-%d"),]
selectNames <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP" ,"CROPDMG","CROPDMGEXP")
data<-select(recentdata, selectNames)

```

##2.3 Regrouping EVTYPE

```{r}
data$EVTYPE[grepl("FLOOD", data$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
data$EVTYPE[grepl("TORNADO", data$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
data$EVTYPE[grepl("TSTM|THUNDERSTORM", data$EVTYPE, ignore.case = TRUE)] <- "TSTM" 
data$EVTYPE[grepl("TROPICAL|STORM", data$EVTYPE, ignore.case = TRUE)] <- "STORM"
data$EVTYPE[grepl("HURRICANE", data$EVTYPE, ignore.case = TRUE)] <- "HURRICANE"  
data$EVTYPE[grepl("ICE|SNOW|FROST|SLEET", data$EVTYPE, ignore.case = TRUE)] <- "SNOW"
data$EVTYPE[grepl("FOG", data$EVTYPE, ignore.case = TRUE)] <- "FOG"
data$EVTYPE[grepl("COLD|WINDCHILL|FREEZE|WINTER", data$EVTYPE, ignore.case = TRUE)] <- "COLD"
data$EVTYPE[grepl("HEAT|WARM|HOT", data$EVTYPE, ignore.case = TRUE)] <- "HEAT"
data$EVTYPE[grepl("CLOUD|FUNNEL", data$EVTYPE, ignore.case = TRUE)] <- "CLOUD" 
data$EVTYPE[grepl("HAIL", data$EVTYPE, ignore.case = TRUE)] <- "HAIL"
data$EVTYPE[grepl("DROUGHT|DRY", data$EVTYPE, ignore.case = TRUE)] <- "DROUGHT"
data$EVTYPE[grepl("LIGHTNING", data$EVTYPE, ignore.case = TRUE)] <- "LIGHTNING"
data$EVTYPE[grepl("FIRE", data$EVTYPE, ignore.case = TRUE)] <- "FIRE" 
data$EVTYPE[grepl("RAIN|SHOWER", data$EVTYPE, ignore.case = TRUE)] <- "RAIN"   
data$EVTYPE[grepl("WATERSPOUT", data$EVTYPE, ignore.case = TRUE)] <- "WATERSPOUT"   
data$EVTYPE[grepl("SURF", data$EVTYPE, ignore.case = TRUE)] <- "SURF"   
data$EVTYPE[grepl("CURRENT", data$EVTYPE, ignore.case = TRUE)] <- "CURRENT"    
data$EVTYPE[grepl("WIND|MICROBURST", data$EVTYPE, ignore.case = TRUE)] <- "WIND" 
data$EVTYPE[grepl("BLIZZARD", data$EVTYPE, ignore.case = TRUE)] <- "BLIZZARD"
data$EVTYPE[grepl("SLIDE", data$EVTYPE, ignore.case = TRUE)] <- "LANDSLIDE"
data$EVTYPE[grepl("DUST", data$EVTYPE, ignore.case = TRUE)] <- "DUST"        
data$EVTYPE<-factor(data$EVTYPE)

```

##2.4 Calculation of Property and Corp Damage
Property damage is indicated by two variables PROPDMG and PROPDMGEXP, same for crop damage. This step firstly converts magnitude characters in PROPDMGEXP and CROPDMGEXP to numeric values and multiplies the values with PROPDMG and CROPDMG respectively.

```{r}
data$PROPDMGEXP<-recode(data$PROPDMGEXP,'K'=1000,'M'=1000000,'B'=1000000000,.default=1)
data$CROPDMGEXP<-recode(data$CROPDMGEXP,'K'=1000,'M'=1000000,'B'=1000000000,.default=1)
data$PROPDMGVALUE <- data$PROPDMG*data$PROPDMGEXP
data$CROPDMGVALUE <- data$CROPDMG*data$CROPDMGEXP

```

#3. Result
##3.1 Most harmful event types to population health
The harm of events to population health is evaluated by 
-fatalities and
- injuries caused collectively.
 
The top 10 most harmful event types are illustrated below.

```{r}

healthdata<-(data %>% group_by(EVTYPE) %>%
               summarise(FATALITIES = sum(FATALITIES),INJURIES = sum(INJURIES) )
               %>% arrange(desc(FATALITIES+INJURIES)))

mostHarm<-healthdata[1:10,]
print(mostHarm)

plotdata<-gather(mostHarm, TYPE, VALUE, FATALITIES:INJURIES)   
ggplot(plotdata, aes(x=reorder(EVTYPE,-VALUE), y=VALUE, fill=TYPE))+
geom_bar(stat="identity")+labs(title="Harmful Events to Population Health", 
                               x="Event Type", y="Count")  

```

 
##3.2 Event types with the greatest economic consequences
The greatest economic consequences are evaluated by 
-property and 
-crop damage collectively.
 
The top 10 event types with greatest economic consequences are shown below.

```{r}
econdata<-(data %>% group_by(EVTYPE) %>% summarise(PROPDMGVALUE =
sum(PROPDMGVALUE),CROPDMGVALUE = sum(CROPDMGVALUE) ) %>% 
arrange(desc(PROPDMGVALUE+CROPDMGVALUE)))


mostEcon<-econdata[1:10,]
plotdata2<-gather(mostEcon, TYPE, VALUE, PROPDMGVALUE:CROPDMGVALUE)  
ggplot(plotdata2, aes(x=reorder(EVTYPE,-VALUE), 
                      y=VALUE, fill=factor(TYPE, 
                      labels=c("crop damage", "property damage"))))+
  geom_bar(stat="identity")+labs(title="Economically Harmful Events",
x="Event Type", y="Count")+guides(fill = guide_legend(title = "Type of damage")) 

```

#4. General Conclusion
By analyzing the data, we can conclude that in the United States, tornadoes are the most damaging event to the health of the population;
Across the United States, floods have the greatest economic consequences.